---
title: "EDA"
output: github_document
date: "2024-10-03"
---

```{r setup, message=FALSE}
library(tidyverse)
library(haven)
library(viridis)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())

```
## initial numeric exploration
```{r}
weather_df %>% 
  ggplot(aes(x = prcp))+
  geom_histogram()
```
```{r}
weather_df |> 
  filter(prcp >= 1000)
```

```{r}
weather_df %>% 
  filter(tmax >= 20, tmax <= 30) %>% 
  ggplot(aes(x = tmin, y = tmax, color = name))+
  geom_point(alpha = 0.75)+
  labs(
    color = "location"
  )+
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )
```
## group_by()

### counting stuff
count the number of observations in each month
count the number of months observed
```{r}
weather_df %>% 
  group_by(name) %>%
  summarize(
    n_obs = n(),
    n_dist = n_distinct(month)
            )
```
group by more than 1 variables
how many observations are there each location each month
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n_obs = n()
  )
```
can summarize more than one thing
```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs_this_month= n(),
    n_days = n_distinct(date)
  )
```
## 2x2 Tables
look at the number of cold and not-cold days in Central Park and Waterhole
```{r}
weather_df %>% 
  drop_na(tmax) %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold"
    )) %>% 
      filter(name != "Molokai_HI") %>% 
   group_by(name, cold) %>% 
  summarize(count = n())
```
```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmax = median(tmax),
    sd_tmax = sd(tmax)
  )
```
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax),
    median_tmax = median(tmax)
  )
```
if you want to summarize multiple columns, use the across() function
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    across(tmin:prcp, mean)
  )
```
creating plots based on the summary
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name))+
  geom_point() + geom_line() +
  theme(legend.position = "bottom")+
  labs(
    color = "Location",
    title = "Monthly Highest Temperature",
    y = "temperature C",
    x = "month"
  )
```
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>% 
  knitr::kable(digits = 1)
  
```
### Grouped mutate
Summarizing collapses groups into single data points. In contrast, using mutate() in conjuntion with group_by() will retain all original data points and add new variables computed within groups.

highest temp in each month in each location
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  mutate(temp_ranking = min_rank(tmax)) %>% 
  filter(temp_ranking < 2)
```
keep the three days with the highest temperature
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  mutate(temp_ranking = min_rank(desc(tmax))) %>% 
  filter(temp_ranking < 4)
```
lag() creates a copy and shift the entries down by 1
```{r}
weather_df %>% 
  mutate(
    lagged_temp = lag(tmax),
    temp_change = tmax - lagged_temp
  ) 
```
```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(temp_change = tmax - lag(tmax)) %>% 
  summarize(
    temp_change_sd = sd(temp_change, na.rm = TRUE)
  )
```

## pulse data
```{r}
pulse_df =
  read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) %>% 
  group_by(visit) %>% 
  summarize(
    mean_bdi = mean(bdi_score, na.rm = TRUE),
    median_bdi = median(bdi_score, na.rm = TRUE)
  )


pulse_df
```





```{r}

pulse_df =
  read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  )
pulse_df %>% 
  group_by(visit) %>% 
  summarize(
    mean_bdi = mean(bdi_score, na.rm =TRUE)
  ) %>% 
  knitr::kable(digits = 1)
```




