---
title: "viz_and_eda"
output: github_document
date: "2024-09-26"
---

```{r}
library(tidyverse)
library(ggridges)
```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```
## Basic scatterplot
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```
## advanced scatterplot
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point()
```

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax))+
  geom_point(aes(color = name), alpha = 0.5, size = 0.8) +
  geom_smooth(se = FALSE)
```

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3, size = .8) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```
```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name, size = prcp))+
  geom_point(alpha = 0.3)+
  geom_smooth(se=FALSE)+
  facet_grid(.~name)

```
```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(
    tmax_f = tmax * (9/5) +32,
    tmin_f = tmin * (9/5) +32
  ) %>% 
  ggplot(aes(x = tmin_f , y = tmax_f))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)
```
## odds and ends
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax))+
  geom_hex()
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax))+
  geom_bin2d()
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax))+
  geom_density2d()
```
Learning Assessment: In the preceding, we set the alpha aesthetic “by hand” instead of mapping it to a variable. This is possible for other aesthetics too. To check your understanding of this point, try to explain why the two lines below don’t produce the same result:
```{r}
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin), color = "blue")
ggplot(weather_df) + geom_point(aes(x = tmax, y = tmin, color = "blue"))
```
In the first attempt, we’re defining the color of the points by hand; in the second attempt, we’re implicitly creating a color variable that has the value blue everywhere; ggplot is then assigning colors according to this variable using the default color scheme.

## univariate plots
```{r}
weather_df %>% 
  ggplot(aes(x = tmax))+
  geom_histogram(aes(fill = name), position = "dodge", binwidth = 1)
```


```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill = name))+
  geom_density(alpha = 0.5, adjust = 0.5)+
  geom_rug()
```

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax))+
  geom_violin(aes(fill = name), alpha = 0.5) +
  stat_summary(fun = "median", color = "blue")
```
```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax))+
  geom_boxplot()
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmax, y = name))+
  geom_density_ridges(scale = 0.85)
```

LA: Make plots that compare precipitation across locations. Try a histogram, a density plot, a boxplot, a violin plot, and a ridgeplot; use aesthetic mappings to make your figure readable.

```{r}
weather_df %>% 
  ggplot(aes(x = prcp, fill = name))+
  geom_density(alpha = 0.3)
```

```{r}
weather_df %>% 
  ggplot(aes(x = prcp, fill = name))+
  geom_boxplot()
```

```{r}
weather_df %>% 
  filter(prcp>10, prcp<1000) %>% 
  ggplot(aes(x = prcp, fill = name))+
  geom_density(alpha = 0.3)
```
```{r}
weather_df %>% 
  filter(prcp>10, prcp<1000) %>% 
  ggplot(aes(x = prcp, y = name))+
  geom_density_ridges(scale = 0.85)
```

##saving and embedding plots
```{r}
weather_df %>% 
  filter(prcp>10, prcp<1000) %>% 
  ggplot(aes(x = prcp, y = name))+
  geom_density_ridges(scale = 0.85)

ggsave("plots/ggp_weather.pdf", width = 8, height = 6)
```
```{r}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```








